Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-07-31 improved_state_space.tex
edits to intro and new bibitem

\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, open population point 
count models, population dynamics, random observer effects, Ricker-logistic model, zero-inflated

@@

\section*{Introduction}
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, 
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

@@

Recently developed state-space models 
have made it possible to
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be overestimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \nonumber \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored, or
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

@@

With respect to the observation error, the Gaussian assumption
implies that false positive detections are
equally likely as false negatives, a pattern inconsistent with most
findings \citep{miller_etal:2011}. 
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as it being present but unavailable for detection,
or to individuals being available but missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models of population dynamics \citep[but
see][]{buckland_etal:2004}.

Other limitations of state-space models, as commonly applied in ecology,
include the failure to admit spatial variation in abundance and
the fact that some parameters are not identifiable
\citep{polansky_etal:2009}. 
Ignoring spatial variation reduces the scope of the
inferences and can lead to underestimation of variances. 
Identifiability problems require 
model constraints or the assertion of additional assumptions. 
Specifically, abundance in
the first year cannot be estimated because there is only one data
point (the count in the first year) that carries any information about
it. To remedy this, practitioners either assume that there is no
observation error in the first year, or they assume the population is
at equilibrium, which may defeat the purpose of many studies of
population dynamics.

>>>>

\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, 
%% open population point count models, 
population dynamics, 
%% random observer effects, 
Ricker-logistic model, zero-inflated

@@

\section*{Introduction}
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, 
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
In applied contexts, population models are used for estimating extinction risk % probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

@@

Recently developed state-space models 
have made it possible to
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a latent %hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be overestimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential. %% Exponential or geometric?
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \nonumber \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; %% Should this be \eta_t?
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian variables: %deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored, or
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

@@

With respect to the observation error, the Gaussian assumption
implies that false positive detections are
equally likely as false negatives, a pattern inconsistent with most
findings \citep{miller_etal:2011}. 
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as it being present but unavailable for detection,
or to individuals being available but missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised to account for %for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models of population dynamics 
%\citep[but see][]{buckland_etal:2004}. 
Of the exceptions that exist, none provide software to facilitate model fitting 
\citep{buckland_etal:2004}.

Other limitations of state-space models, as commonly applied in ecology,
include the failure to admit spatial variation in abundance and
the fact that some parameters are not identifiable
\citep{polansky_etal:2009}. 
Ignoring spatial variation reduces the scope of the
inferences and reduces the information about the parameters of interest \citep{dennis_etal:2010}.
%% Identifiability problems require 
%% model constraints or the assertion of additional assumptions. 
%% Specifically, 
Even more problematic is the issue that without some form of replication in the first year, 
abundance cannot be estimated without making strong assumptions about 
the initial distribution. 
%% because there is only one data
%% point (the count in the first year) that carries any information about
%%it. To remedy this, 
For example, practitioners either assume that there is no
observation error in the first year, or they assume the population is
at equilibrium, which may defeat the purpose of many studies of
population dynamics.
